---
title: "COV19 in North Rhein-Westphalia"
output:
  rmarkdown::html_document
---

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = TRUE,
  message = TRUE,
  comment = "#>",
  width = "100%",
  fig.path = "docs/"
)
```

This is all just a direct translation of the analyses of [Tim
Churches](https://timchurches.github.io), primarily from [this blog
post](https://timchurches.github.io/blog/posts/2020-02-18-analysing-covid-19-2019-ncov-outbreak-data-with-r-part-1).
See also [part 2 of that blog
post](https://timchurches.github.io/blog/posts/2020-03-01-analysing-covid-19-2019-ncov-outbreak-data-with-r-part-2/),
and a higher-level description on the [RStudio Community
Blog](https://rviews.rstudio.com/2020/03/05/covid-19-epidemiology-with-r). All
gratitude and credit to Tim Churches.

```{r libraries, message = FALSE}
library (lubridate)
library (tidyr)
library (ggplot2)
theme_set (theme_light ())
library (dplyr)
# Specific libraries for epidemic analyses:
library (incidence)
library (epitrix)
library (distcrete)
library (EpiEstim)
```


The raw data in this case are from the wikipedia site for
[2020_coronavirus_outbreak_in_Germany](https://en.wikipedia.org/wiki/2020_coronavirus_outbreak_in_Germany),
which contains the most up-to-date information. This site will be updated
daily, at least until things are clearly under control (last update `r
today()`). The raw data as presented in wikipedia are presented here:

```{r raw-data}
x <- rbind (c ("Baden-Württemberg", 0, 1, 3, 6, 10, 14, 15, 19, 26, 44, 65, 91, 170),
            c ("Bavaria", 14, 14, 14, 14, 15, 15, 19, 25, 35, 48, 52, 79, 134),
            c ("Berlin", 0, 0, 0, 0, 0, 0, 0, 1, 3, 6, 9, 15, 28),
            c ("Brandenburg", 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 2, 2),
            c ("Bremen", 0, 0, 0, 0, 0, 0, 1, 1, 1, 2, 3, 4, 4),
            c ("Hamburg", 0, 0, 0, 0, 0, 0, 1, 1, 2, 2, 8, 11),
            c ("Hessen", 0, 0, 0, 0, 0, 3, 8, 10, 10, 12, 14, 15, 17),
            c ("Mecklenburg-Vorpommern", 0, 0, 0, 0, 0, 0, 0, 0, 0, 3, 4, 5, 5),
            c ("Lower Saxony", 0, 0, 0, 0, 0, 0, 1, 1, 1, 4, 10, 18, 19),
            c ("North Rhine-Westphalia", 0, 1, 2, 4, 25, 30, 66, 86, 101, 111, 175, 281, 373),
            c ("Rhineland-Palatinate", 0, 0, 0, 0, 0, 1, 2, 2, 2, 2, 7, 8, 13),
            c ("Saarland", 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 3),
            c ("Saxony", 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 4),
            c ("Saxony-Anhalt", 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
            c ("Schleswig-Holstein", 0, 0, 0, 0, 1, 1, 2, 2, 2, 2, 3, 7, 8),
            c ("Thuringia", 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 2))
state <- x [, 1]
x <- data.frame (apply (x [, 2:ncol (x)], 2, function (i) as.integer (i)))
ndays <- ncol (x)
names (x) <- paste0 (ymd ("2020-02-24") + days (seq (ndays) - 1))
x <- cbind (state = state, x)
```

Current trajectories for each state of Germany look like this:

```{r germany-plot, message = FALSE}
xl <- pivot_longer (x, col = names (x) [-1])
xl$name <- ymd (xl$name)
xl <- rename (xl, date = name, cases = value)
ggplot (xl, aes (x = date, y = cases, colour = state)) +
    geom_line () +
    geom_point ()
```

Tim's functions for fitting the main "SIR" (Susceptible, Infectious, Recovered)
model are as follows.

```{r main-model}
SIR <- function(time, state, parameters) {
    par <- as.list(c(state, parameters))
    with(par, {
             dS <- -beta * I * S/N
             dI <- beta * I * S/N - gamma * I
             dR <- gamma * I
             list(c(dS, dI, dR))
            })
}

# put the daily cumulative incidence numbers into a vector called Infected
state <- "North Rhine-Westphalia"
sir_start_date <- min (xl$date)
sir_latest_date <- max (xl$date)
Infected <- as.integer (x [x$state == state, 2:ncol (x)])
index <- min (which (Infected > 0)):length (Infected)
dates <- sir_start_date + days ((1:ndays - 1) [index])
Infected <- Infected [index]
Day <- seq_along (Infected)

# now specify initial values for S, I and R
N <- 17.9e6 # current wikipedia population of NRW
init <- c(S = N - Infected[1], I = Infected[1], R = 0)

# define a function to calculate the residual sum of squares
# (RSS), passing in parameters beta and gamma that are to be
# optimised for the best fit to the incidence data
RSS <- function(parameters) {
    names(parameters) <- c("beta", "gamma")
    out <- deSolve::ode (y = init, times = Day, func = SIR, parms = parameters)
    fit <- out [, 3]
    sum ((Infected - fit) ^ 2)
}

# now find the values of beta and gamma that give the
# smallest RSS, which represents the best fit to the data.
# Start with values of 0.5 for each, and constrain them to
# the interval 0 to 1.0
Opt <- optim (c (0.5, 0.5), RSS, method = "L-BFGS-B",
              lower = c(0, 0), upper = c(1, 1))

# check for convergence
Opt$message

Opt_par <- setNames(Opt$par, c("beta", "gamma"))
Opt_par
```
Those parameters are used to determine the basic reproductive rate of the
virus.

```{r R0-estimate}
# get the fitted values from our SIR model
fitted_cumulative_incidence <- data.frame (deSolve::ode (y = init,
                                                         times = Day, 
                                                         func = SIR,
                                                         parms = Opt_par))
fitted_cumulative_incidence$observed <- Infected
fitted_cumulative_incidence$date <- dates

# plot the data
main <- paste0 ("COVID-19 fitted vs observed cumulative incidence, ", state)
fitted_cumulative_incidence %>%
    ggplot(aes(x = date)) + geom_line(aes(y = I), colour = "red") + 
    geom_point(aes(y = observed), colour = "orange") + 
    labs(y = "Cumulative incidence", title = main,
         subtitle = "(red=fitted incidence from SIR model, orange=observed incidence)")

R0 <- Opt_par [names (Opt_par) == "beta"] / Opt_par [names (Opt_par) == "gamma"]
message ("Basic reproductive rate = ", formatC (R0, format = "f", digits = 2))
```

That initial estimate is *extremely* high (compared, for example, with Tim's
estimate of around 2.0 for the initial cases from Hubei Province, China, the
origin of the current outbreak. The fitted model also does not appear to fit
the observed data particularly well in this case, so that estimate ought not at
the current point be taken as accurate. We also currently only have 11 days of
data for North Rhein-Westphalia, compared with 22 days used by Tim for his
analysis of Hubei Province. The summary message at present is that these
analyses appear to be currently quite inaccurate, but will hopefully increase
in accuracy in the coming days as more data become available.

## prediction

Tim's code to use the model derived above to generate predicted values follows:

```{r prediction}
t <- 1:70
# get the fitted values from our SIR model
fitted_cumulative_incidence <- data.frame(deSolve::ode(y = init, times = t, 
                                                       func = SIR, parms = Opt_par))
fitted_cumulative_incidence$date <- ymd (sir_start_date) + days (t - 1)
fitted_cumulative_incidence$observed <- c (Infected, rep (NA, nrow (fitted_cumulative_incidence) - length (Infected)))

sir <- pivot_longer (fitted_cumulative_incidence, cols = c (I, S, R)) %>%
    rename (type = name)
types <- c ("Infectious", "Susceptible", "Recovered")
sir$type <- types [match (sir$type, c ("I", "S", "R"))]
ggplot (sir, aes (x = date, y = value, colour = type)) +
    geom_line () +
    geom_point (aes (y = observed), colour = "orange") +
    scale_y_log10 () +
    labs (y = "Person", title = main)
```

That model predicts a peak date of 
```{r}
dmax <- fitted_cumulative_incidence$date [which.max (fitted_cumulative_incidence$I)]
dmax
```
so we've still got `r as.integer (dmax - today ())` days to get through.



## Reproduction Number

Calculating the reproductive rate requires first converting the incidence data
into a simple vector of dates (via the `tidyr::uncount()` function), in which
each entry represents a single case. The very large population of North
Rhein-Westphalia (`r format (N, big.mark = ",")`) translates to a vector too
long to be constructed here, so the incidence rates are first divided by 10,
before being subsequently re-scaled. Note throughout this that the preceding
estimate of reproductive rate seems erroneous given the current data from North
Rhein-Westphalia, and so the following analysis should also be considered for
the moment to be inaccurate.

```{r incidence}
x <- fitted_cumulative_incidence %>%
    select (date, I) %>%
    transform (I = round (I / 10)) %>%
    uncount (I)
incidence_obj <- incidence (x$date)
incidence_peak <- find_peak (incidence_obj)
# Then restore the counts to their original scale
incidence_obj$counts <- incidence_obj$counts * 10

plot (incidence_obj) +
    geom_vline(xintercept = incidence_peak, col = "red", lty = 2) +
    labs(title = paste0 ("Daily incidence of lab-confirmed cases, ", state),
         subtitle = "(red line indicates date of peak incidence)")
```

The next step is to fit a model to the two phases of the epidemic's dynamic:
growth and decay.
```{r incidence-fit}
incidence_fit <- incidence::fit(incidence_obj, split = incidence_peak)

plot(incidence_obj) %>%
    add_incidence_fit(incidence_fit) + 
    labs(title = "Observed and modelled incidence of COVID-19 cases", 
         subtitle = state)
```

Distributions for reproductive rates can then be fitted using the
`incidence_fit` model:

```{r distcrete}
mu <- 7.5  # days
sigma <- 3.4  # days
param <- gamma_mucv2shapescale(mu, sigma/mu)

w <- distcrete("gamma", interval = 1,
               shape = param$shape, scale = param$scale, w = 0)

growth_R0 <- lm2R0_sample(incidence_fit$before$model, w)
hist(growth_R0, col = "grey", border = "white", main = "Distribution of R0")

decay_R0 <- lm2R0_sample(incidence_fit$after$model, w)
hist(decay_R0, col = "grey", border = "white", main = "Distribution of R0")
```

The former is way too high, because of the apparent initial over-estimate of
R0, and the latter also consequently represents a rate of decay higher than it
might actually likely be.

### changes in R

Tim then proceeds to model likely changes in reproductive rate through time
using the following code.

```{r changes-in-R}
# custom results plotting function to avoid the ugly
# TableGrob messages returned by the plotting function in the
# EpiEstim package
plot_Ri <- function(estimate_R_obj) {
    p_I <- plot(estimate_R_obj, "incid", add_imported_cases = TRUE)  # plots the incidence
    p_SI <- plot(estimate_R_obj, "SI")  # plots the serial interval distribution
    p_Ri <- plot(estimate_R_obj, "R")
    return(gridExtra::grid.arrange(p_I, p_SI, p_Ri, ncol = 1))
}

confirmed_cases <- Infected
res_parametric_si <- estimate_R(confirmed_cases, method = "parametric_si",
                                config = make_config(list(mean_si = 7.5, 
                                                          std_si = 3.4)))

plot_Ri(res_parametric_si)
```

And at least that shows a very pronounced decrease in `R`, which is
encouraging. He then fits a more sophisticated model, effectively presuming
reproductive rates themselves may vary on a case-to-case basis, and so are
drawn from an underlying generative distribution, parameterised as follows:

```{r changes-in-R2}
config <- make_config(list(mean_si = 7.5, std_mean_si = 2,
                      min_mean_si = 1, max_mean_si = 8.4,
                      std_si = 3.4, std_std_si = 1,
                      min_std_si = 0.5, max_std_si = 4, n1 = 1000, n2 = 1000))
res_parametric_si <- estimate_R(confirmed_cases, method = "parametric_si",
                                config = config)
plot_Ri(res_parametric_si)
```

Those values are results ought not be taken as accurate at present, and
importantly use Tim's estimates for mean and standard deviation of the serial
interval (SI), which is the time between the onset of symptoms in any one case,
and the onset of symptoms in secondary cases caused by that first infection.

When more NRW data become available, can then continue the analysis from
https://timchurches.github.io/blog/posts/2020-02-18-analysing-covid-19-2019-ncov-outbreak-data-with-r-part-1/#estimating-changes-in-the-effective-reproduction-number



